---
header-includes:
- \usepackage{pdflscape}
- \usepackage{colortbl}
- \usepackage{texshade}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
output: 
  pdf_document: 
    keep_tex: yes
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE, warning = FALSE)
options(knitr.table.format = "latex") 
library(ggplot2)
library(ggforce)
library(knitr)
library(dplyr)
library(tidyr)
library(kableExtra)
library(merf)

####
# Replace with local project directory
####
projectDir <- "/home/kevin/projects/viralTransmission/"

rawData <- read.table(file=paste0(projectDir, "viralTransmissionData.tsv"),header = TRUE,as.is = TRUE,fill = TRUE, sep = "\t", quote = "") %>%
  arrange(Common.Name)

data <- rawData %>% filter(unclear_transmission==FALSE) %>%
  mutate(Lipid.Envelope=as.logical(Lipid.Envelope),
         Spherical=as.logical(Spherical))

printHumanHeatmap = TRUE
printHeatmap = FALSE
printByHost = TRUE
printTestFigures = TRUE
printTests = FALSE

transmissionVars <- c("Fecal.Oral","Arbovirus","Inhalation.Aerosols","Inhalation.Dust","Sexual","Eating","Oral.Bloodstream","Breastfeeding","Maternal.Fetal","Germ.line","Blood.Products","Contact.Skin.or.Eye")
transmissionCols <- which(colnames(data) %in% transmissionVars)

attributeVars <- c("NucAcid","Spherical","Filamentous","Pleomorphic","Bullet.form","Lipid.Envelope")
attributeCols <- which(colnames(data) %in% attributeVars)

humanData <- data %>% filter(Common.Name == 'Human')
nonhumanData <- data %>% filter(Common.Name != 'Human')

vfData <- data %>%
  group_by(Host.species, Common.Name, Virus.family, NucAcid) %>%
  summarize_if(is.logical, function(X) { if(sum(X) > 0) { return(TRUE) } else { return(FALSE)}}) %>%
  ungroup() %>%
  as.data.frame()

famTransmissionCols <- which(colnames(vfData) %in% transmissionVars)
famAttributeCols <- which(colnames(vfData) %in% attributeVars)

vfDataH <- data %>% filter(Common.Name == "Human") %>%
  group_by(Host.species, Common.Name, Virus.family, NucAcid) %>%
  summarize_if(is.logical, function(X) { if(sum(X) > 0) { return(TRUE) } else { return(FALSE)}}) %>%
  ungroup() %>%
  as.data.frame()

famTransmissionColsH <- which(colnames(vfDataH) %in% transmissionVars)
famAttributeColsH <- which(colnames(vfDataH) %in% attributeVars)

vfDataNH <- data %>% filter(Common.Name != "Human") %>%
  group_by(Host.species, Common.Name, Virus.family, NucAcid) %>%
  summarize_if(is.logical, function(X) { if(sum(X) > 0) { return(TRUE) } else { return(FALSE)}}) %>%
  ungroup() %>%
  arrange(Common.Name) %>%
  as.data.frame()

famTransmissionColsNH <- which(colnames(vfDataNH) %in% transmissionVars)
famAttributeColsNH <- which(colnames(vfDataNH) %in% attributeVars)

```

# Overview

This report serves to investigate the relationship between various modes of transmission of viruses causing disease in humans and other attributes of those viruses (physical structure, classification, etc.).  This report includes metadata for `r nrow(rawData)` individual viral species.

# Host Species

Our data included viruses from `r nrow(rawData %>% distinct(Common.Name))` host species.  Our analyses below include a global view as well as a look at the viruses for each individual host species.

# Initial Filter

Our data set of `r nrow(rawData)` individual viral species included `r nrow(rawData %>% filter(unclear_transmission==TRUE))` species with an unclear mechanism of transmission and thus were excluded from our analysis.

# Statistical Methods

We looked at the relationship between `r length(transmissionVars)`* different modes of transmission against `r length(attributeVars)` other viral attributes.  For each mode-attribute pair, we performed a Fisher's exact test to see if there is a non-random association between the variables.  The initial p-values from these tests were adjusted using the Benjamini-Hochberg correction method.


*Our data initially separated modes of transmission by mosquito, tick, and other arthropods, but for statistical analyses we grouped these into one test for arboviruses.

```{r virus.counts.by.host, echo=FALSE, eval=TRUE, results='asis', fig.height=3}

cat(paste0("\n\n# Virus count by host species\n"))

rawData %>% group_by(Common.Name) %>%
  summarize(count=n()) %>%
  kable(col.names=c("Host","Virus count"))

```

```{r figures, echo=FALSE, eval=printTestFigures, results='asis', fig.height=3}

cat("\n\n\\pagebreak\n")
cat(paste0("\n\n# Analysis for all host species\n"))

pValues <- NULL
odds <- NULL
confIntLow <- NULL
confIntHigh <- NULL
iVals <- NULL
jVals <- NULL

for(i in attributeCols) {
  if(length(unique(data[,i]))==2) {
    for(j in transmissionCols) {
      if(length(unique(data[,j]))==2) {
        tbl <- table(data[,i],data[,j])
        ft <- fisher.test(tbl)
        pValues <- append(pValues,ft$p.value)
        iVals <- append(iVals, i)
        jVals <- append(jVals, j)
        odds <- append(odds, ft$estimate)
        confIntLow <- append(confIntLow, ft$conf.int[1])
        confIntHigh <- append(confIntHigh, ft$conf.int[2])
      }
    }
  }
}

pValuesAdj <- p.adjust(pValues, method = 'BH')
names(pValuesAdj) <- pValues
pThreshold <- 0.05

data.frame(x=sort(pValues),
           y=sort(pValuesAdj)) %>%
  mutate(significance=if_else(x < pThreshold, "significant before correction", "not significant")) %>%
  mutate(significance=if_else(y < pThreshold, "significant", significance)) %>%
  ggplot(aes(x=x,y=y, color=significance)) +
  geom_point() +
  scale_color_manual(values=c("significant"="red","not significant"="black", "significant before correction"="ForestGreen")) +
  geom_hline(yintercept = pThreshold) + geom_vline(xintercept = pThreshold) +
  xlab("Significance ") + ylab("p-value")


print(data.frame(pval=1/pValuesAdj,
                 odds=odds,
                 cil=confIntLow,
                 cih=confIntHigh,
                 i=iVals,
                 j=jVals,
                 label=paste0(colnames(data)[iVals], " vs ",colnames(data)[jVals]),
                 stringsAsFactors = FALSE) %>%
        filter(pValuesAdj < 0.05) %>%
        mutate(maxcih=max(cih[is.finite(cih)])) %>%
        mutate(odds=if_else(is.finite(odds),odds,maxcih),
               cih=if_else(is.finite(cih),cih,maxcih)) %>%
        arrange(-pval, odds, -cih, cil) %>%
        arrange(pval, -odds, -cih, cil) %>%
        mutate(label=factor(label,levels=label)) %>%
        ggplot(aes(x=label,y=odds)) +
        geom_point(stat="identity") +
        geom_segment(aes(x=label,xend=label,y=cil,yend=cih)) +
        geom_hline(yintercept = 1, color="red") +
        ylab("Odds ratio (+ confidence interval)") + xlab("Test performed") +
        scale_y_log10() +
        coord_flip())

```

```{r fisher.tests, echo=FALSE, eval=printTests, results='asis', fig.height=3}


for(i in attributeCols) {
  attribute <- colnames(data)[i]
  cat("\n\n\\pagebreak\n")
  if(length(unique(data[,i]))==2) {
    cat(paste0("\n\n# Comparison of attribute '", attribute, "'"))
    for(j in transmissionCols) {
      transmission <- colnames(data)[j]
      cat(paste0("\n\n### Transmission mode '", transmission, "' vs attribute '", attribute, "'\n"))
      tbl <- table(data[,i],data[,j])
      df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
        mutate(Var1=paste0(attribute, " = ", Var1))
      colnames(df) <- c("",transmission,"other")
      print(kable(df))
      ft <- fisher.test(tbl)
      pValue <- ft$p.value
      pValueAdj <- pValuesAdj[as.character(pValue)]
      color <- if_else(pValueAdj < pThreshold,"red",if_else(pValue < pThreshold, "ForestGreen", "black"))
      cat(paste0("\\textcolor{",color,"}{p-value = ",round(pValue,digits = 5),"}\\newline"))
      cat(paste0("\\textcolor{",color,"}{adjusted p-value = ",round(pValueAdj,digits = 5),"}"))
    }
  } else {
    for(value in unique(data[,i])) {
      cat(paste0("\n\n# Comparison of value '", value, "' for attribute '", attribute,"'"))
      for(j in transmissionCols) {
        transmission <- colnames(data)[j]
        cat(paste0("\n\n### Transmission mode '", transmission, "' vs value '", value, "'\n"))
        tbl <- table(data[,i]==value,data[,j])
        df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
          mutate_at(vars("Var1"),funs(if_else(Var1=="TRUE",paste0(attribute, " = ", value),paste0(attribute, " !=  ", value))))
        colnames(df) <- c("",transmission,"other")
        print(kable(df))
        ft <- fisher.test(tbl)
        color <- if_else(ft$p.value < pThreshold,"red","black")
        cat(paste0("\\textcolor{",color,"}{p-value = ",round(ft$p.value,digits = 5),"}"))
      }
    }
  }
}

```


```{r figures.human, echo=FALSE, eval=printTestFigures, results='asis', fig.height=3}

cat("\n\n\\pagebreak\n")
cat(paste0("\n\n# Analysis for humans\n"))

pValues <- NULL
odds <- NULL
confIntLow <- NULL
confIntHigh <- NULL
iVals <- NULL
jVals <- NULL

for(i in attributeCols) {
  if(length(unique(humanData[,i]))==2) {
    for(j in transmissionCols) {
      if(length(unique(humanData[,j]))==2) {
        tbl <- table(humanData[,i],humanData[,j])
        ft <- fisher.test(tbl)
        pValues <- append(pValues,ft$p.value)
        iVals <- append(iVals, i)
        jVals <- append(jVals, j)
        odds <- append(odds, ft$estimate)
        confIntLow <- append(confIntLow, ft$conf.int[1])
        confIntHigh <- append(confIntHigh, ft$conf.int[2])
      }
    }
  }
}

pValuesAdj <- p.adjust(pValues, method = 'BH')
names(pValuesAdj) <- pValues
pThreshold <- 0.05
  
print(data.frame(x=sort(pValues),
           y=sort(pValuesAdj)) %>%
  mutate(significance=if_else(x < pThreshold, "significant before correction", "not significant")) %>%
  mutate(significance=if_else(y < pThreshold, "significant", significance)) %>%
  ggplot(aes(x=x,y=y, color=significance)) +
  geom_point() +
  scale_color_manual(values=c("significant"="red","not significant"="black", "significant before correction"="ForestGreen")) +
  geom_hline(yintercept = pThreshold) + geom_vline(xintercept = pThreshold) +
  xlab("Significance ") + ylab("p-value"))

print(data.frame(pval=1/pValuesAdj,
                 odds=odds,
                 cil=confIntLow,
                 cih=confIntHigh,
                 i=iVals,
                 j=jVals,
                 label=paste0(colnames(data)[iVals], " vs ",colnames(data)[jVals]),
                 stringsAsFactors = FALSE) %>%
        filter(pValuesAdj < 0.05) %>%
        mutate(maxcih=max(cih[is.finite(cih)])) %>%
        mutate(odds=if_else(is.finite(odds),odds,maxcih),
               cih=if_else(is.finite(cih),cih,maxcih)) %>%
        arrange(-pval, odds, -cih, cil) %>%
        arrange(pval, -odds, -cih, cil) %>%
        mutate(label=factor(label,levels=label)) %>%
        ggplot(aes(x=label,y=odds)) +
        geom_point(stat="identity") +
        geom_segment(aes(x=label,xend=label,y=cil,yend=cih)) +
        geom_hline(yintercept = 1, color="red") +
        ylab("Odds ratio (+ confidence interval)") + xlab("Test performed") +
        scale_y_log10() +
        coord_flip())

```

```{r fisher.tests.human, echo=FALSE, eval=printTests, results='asis', fig.height=3}

for(i in attributeCols) {
  attribute <- colnames(humanData)[i]
  cat("\n\n\\pagebreak\n")
  if(length(unique(humanData[,i]))==1) {
    next
  }
  if(length(unique(humanData[,i]))==2) {
    cat(paste0("\n\n# Comparison of attribute '", attribute, "' in humans"))
    for(j in transmissionCols) {
      if(length(unique(humanData[,j]))==1) {
        next
      }
      transmission <- colnames(humanData)[j]
      cat(paste0("\n\n### Transmission mode '", transmission, "' vs attribute '", attribute, "' in human\n"))
      tbl <- table(humanData[,i],humanData[,j])
      df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
        mutate(Var1=paste0(attribute, " = ", Var1))
      colnames(df) <- c("",transmission,"other")
      print(kable(df))
      ft <- fisher.test(tbl)
      pValue <- ft$p.value
      pValueAdj <- pValuesAdj[as.character(pValue)]
      color <- if_else(pValueAdj < pThreshold,"red",if_else(pValue < pThreshold, "ForestGreen", "black"))
      cat(paste0("\\textcolor{",color,"}{p-value = ",round(pValue,digits = 5),"}\\newline"))
      cat(paste0("\\textcolor{",color,"}{adjusted p-value = ",round(pValueAdj,digits = 5),"}"))
    }
  } else {
    for(value in unique(humanData[,i])) {
      if(length(unique(humanData[,j]))==1) {
        next
      }
      cat(paste0("\n\n### Transmission mode '", transmission, "' vs attribute '", attribute, "' in humans\n"))
      for(j in transmissionCols) {
        transmission <- colnames(humanData)[j]
        cat(paste0("\n\n### Transmission mode '", transmission, "' vs value '", value, "'\n"))
        tbl <- table(humanData[,i]==value,humanData[,j])
        df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
          mutate_at(vars("Var1"),funs(if_else(Var1=="TRUE",paste0(attribute, " = ", value),paste0(attribute, " !=  ", value))))
        colnames(df) <- c("",transmission,"other")
        print(kable(df))
        ft <- fisher.test(tbl)
        color <- if_else(ft$p.value < pThreshold,"red","black")
        cat(paste0("\\textcolor{",color,"}{p-value = ",round(ft$p.value,digits = 5),"}"))
      }
    }
  }
}

```

```{r figures.nonhuman, echo=FALSE, eval=printTestFigures, results='asis', fig.height=3}

cat("\n\n\\pagebreak\n")
cat(paste0("\n\n# Analysis for all non-human host species\n"))

pValues <- NULL
odds <- NULL
confIntLow <- NULL
confIntHigh <- NULL
iVals <- NULL
jVals <- NULL

for(i in attributeCols) {
  if(length(unique(nonhumanData[,i]))==2) {
    for(j in transmissionCols) {
      if(length(unique(nonhumanData[,j]))==2) {
        tbl <- table(nonhumanData[,i],nonhumanData[,j])
        ft <- fisher.test(tbl)
        pValues <- append(pValues,ft$p.value)
        iVals <- append(iVals, i)
        jVals <- append(jVals, j)
        odds <- append(odds, ft$estimate)
        confIntLow <- append(confIntLow, ft$conf.int[1])
        confIntHigh <- append(confIntHigh, ft$conf.int[2])
      }
    }
  }
}

pValuesAdj <- p.adjust(pValues, method = 'BH')
names(pValuesAdj) <- pValues
pThreshold <- 0.05
  
print(data.frame(x=sort(pValues),
           y=sort(pValuesAdj)) %>%
  mutate(significance=if_else(x < pThreshold, "significant before correction", "not significant")) %>%
  mutate(significance=if_else(y < pThreshold, "significant", significance)) %>%
  ggplot(aes(x=x,y=y, color=significance)) +
  geom_point() +
  scale_color_manual(values=c("significant"="red","not significant"="black", "significant before correction"="ForestGreen")) +
  geom_hline(yintercept = pThreshold) + geom_vline(xintercept = pThreshold) +
  xlab("Significance ") + ylab("p-value"))

print(data.frame(pval=1/pValuesAdj,
                 odds=odds,
                 cil=confIntLow,
                 cih=confIntHigh,
                 i=iVals,
                 j=jVals,
                 label=paste0(colnames(data)[iVals], " vs ",colnames(data)[jVals]),
                 stringsAsFactors = FALSE) %>%
        filter(pValuesAdj < 0.05) %>%
        mutate(maxcih=max(cih[is.finite(cih)])) %>%
        mutate(odds=if_else(is.finite(odds),odds,maxcih),
               cih=if_else(is.finite(cih),cih,maxcih)) %>%
        arrange(-pval, odds, -cih, cil) %>%
        arrange(pval, -odds, -cih, cil) %>%
        mutate(label=factor(label,levels=label)) %>%
        ggplot(aes(x=label,y=odds)) +
        geom_point(stat="identity") +
        geom_segment(aes(x=label,xend=label,y=cil,yend=cih)) +
        geom_hline(yintercept = 1, color="red") +
        ylab("Odds ratio (+ confidence interval)") + xlab("Test performed") +
        scale_y_log10() +
        coord_flip())

```

```{r fisher.tests.nonhuman, echo=FALSE, eval=printTests, results='asis', fig.height=3}


for(i in attributeCols) {
  attribute <- colnames(nonhumanData)[i]
  cat("\n\n\\pagebreak\n")
  if(length(unique(nonhumanData[,i]))==1) {
    next
  }
  if(length(unique(nonhumanData[,i]))==2) {
    cat(paste0("\n\n# Comparison of attribute '", attribute, "' for non-human hosts"))
    for(j in transmissionCols) {
      if(length(unique(nonhumanData[,j]))==1) {
        next
      }
      transmission <- colnames(nonhumanData)[j]
      cat(paste0("\n\n### Transmission mode '", transmission, "' vs attribute '", attribute, "' in non-human hosts\n"))
      tbl <- table(nonhumanData[,i],nonhumanData[,j])
      df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
        mutate(Var1=paste0(attribute, " = ", Var1))
      colnames(df) <- c("",transmission,"other")
      print(kable(df))
      ft <- fisher.test(tbl)
      pValue <- ft$p.value
      pValueAdj <- pValuesAdj[as.character(pValue)]
      color <- if_else(pValueAdj < pThreshold,"red",if_else(pValue < pThreshold, "ForestGreen", "black"))
      cat(paste0("\\textcolor{",color,"}{p-value = ",round(pValue,digits = 5),"}\\newline"))
      cat(paste0("\\textcolor{",color,"}{adjusted p-value = ",round(pValueAdj,digits = 5),"}"))
    }
  } else {
    for(value in unique(nonhumanData[,i])) {
      if(length(unique(nonhumanData[,j]))==1) {
        next
      }
      cat(paste0("\n\n### Transmission mode '", transmission, "' vs attribute '", attribute, "' in non-human hosts\n"))
      for(j in transmissionCols) {
        transmission <- colnames(nonhumanData)[j]
        cat(paste0("\n\n### Transmission mode '", transmission, "' vs value '", value, "'\n"))
        tbl <- table(nonhumanData[,i]==value,nonhumanData[,j])
        df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
          mutate_at(vars("Var1"),funs(if_else(Var1=="TRUE",paste0(attribute, " = ", value),paste0(attribute, " !=  ", value))))
        colnames(df) <- c("",transmission,"other")
        print(kable(df))
        ft <- fisher.test(tbl)
        color <- if_else(ft$p.value < pThreshold,"red","black")
        cat(paste0("\\textcolor{",color,"}{p-value = ",round(ft$p.value,digits = 5),"}"))
      }
    }
  }
}

```

```{r host.fisher.tests, echo=FALSE, eval=printByHost, results='asis', fig.height=3}
for(host in unique(data$Common.Name)) {
  cat("\n\n\\pagebreak\n")
  cat(paste0("\n\n# Analysis for host '", host, "'\n"))
  
  hostData <- data %>% filter(Common.Name==host)
  
  pValues <- NULL
  odds <- NULL
  confIntLow <- NULL
  confIntHigh <- NULL
  iVals <- NULL
  jVals <- NULL
  
  for(i in attributeCols) {
    if(length(unique(hostData[,i]))==2) {
      for(j in transmissionCols) {
        if(length(unique(hostData[,j]))==2) {
          tbl <- table(hostData[,i],hostData[,j])
          ft <- fisher.test(tbl)
          pValues <- append(pValues,ft$p.value)
          iVals <- append(iVals, i)
          jVals <- append(jVals, j)
          odds <- append(odds, ft$estimate)
          confIntLow <- append(confIntLow, ft$conf.int[1])
          confIntHigh <- append(confIntHigh, ft$conf.int[2])
        }
      }
    }
  }
  
  pValuesAdj <- p.adjust(pValues, method = 'BH')
  names(pValuesAdj) <- pValues
  pThreshold <- 0.05
  
  if(printTestFigures) {
    print(data.frame(x=sort(pValues),
               y=sort(pValuesAdj)) %>%
      mutate(significance=if_else(x < pThreshold, "significant before correction", "not significant")) %>%
      mutate(significance=if_else(y < pThreshold, "significant", significance)) %>%
      ggplot(aes(x=x,y=y, color=significance)) +
      geom_point() +
      scale_color_manual(values=c("significant"="red","not significant"="black", "significant before correction"="ForestGreen")) +
      geom_hline(yintercept = pThreshold) + geom_vline(xintercept = pThreshold) +
      xlab("Significance ") + ylab("p-value"))
  
    print(data.frame(pval=1/pValuesAdj,
                     odds=odds,
                     cil=confIntLow,
                     cih=confIntHigh,
                     i=iVals,
                     j=jVals,
                     label=paste0(colnames(data)[iVals], " vs ",colnames(data)[jVals]),
                     stringsAsFactors = FALSE) %>%
            filter(pValuesAdj < 0.05) %>%
            mutate(maxcih=max(cih[is.finite(cih)])) %>%
            mutate(odds=if_else(is.finite(odds),odds,maxcih),
                   cih=if_else(is.finite(cih),cih,maxcih)) %>%
            arrange(-pval, odds, -cih, cil) %>%
            arrange(pval, -odds, -cih, cil) %>%
            mutate(label=factor(label,levels=label)) %>%
            ggplot(aes(x=label,y=odds)) +
            geom_point(stat="identity") +
            geom_segment(aes(x=label,xend=label,y=cil,yend=cih)) +
            geom_hline(yintercept = 1, color="red") +
            ylab("Odds ratio (+ confidence interval)") + xlab("Test performed") +
            scale_y_log10() +
            coord_flip())
  }
  
  if(printTests) {
    for(i in attributeCols) {
      attribute <- colnames(hostData)[i]
      cat("\n\n\\pagebreak\n")
      if(length(unique(hostData[,i]))==2) {
        cat(paste0("\n\n# Comparison of attribute '", attribute, "' for host ", host))
        for(j in transmissionCols) {
          if(length(unique(hostData[,j]))==2) {
            transmission <- colnames(hostData)[j]
            cat(paste0("\n\n### Transmission mode '", transmission, "' vs attribute '", attribute, "' in ", host, "\n"))
            tbl <- table(hostData[,i],hostData[,j])
            df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
              mutate(Var1=paste0(attribute, " = ", Var1))
            colnames(df) <- c("",transmission,"other")
            print(kable(df))
            ft <- fisher.test(tbl)
            pValue <- ft$p.value
            pValueAdj <- pValuesAdj[as.character(pValue)]
            color <- if_else(pValueAdj < pThreshold,"red",if_else(pValue < pThreshold, "ForestGreen", "black"))
            cat(paste0("\\textcolor{",color,"}{p-value = ",round(pValue,digits = 5),"}\\newline"))
            cat(paste0("\\textcolor{",color,"}{adjusted p-value = ",round(pValueAdj,digits = 5),"}"))
          }
        }
      } 
    }
  }
}

```

```{r figures.viralFamilies, echo=FALSE, eval=printTestFigures, results='asis', fig.height=3}

cat("\n\n\\pagebreak\n")
cat(paste0("\n\n# Analysis for viral families in all host species\n"))


pValues <- NULL
odds <- NULL
confIntLow <- NULL
confIntHigh <- NULL
iVals <- NULL
jVals <- NULL

for(i in famAttributeCols) {
  if(length(unique(vfData[,i]))==2) {
    for(j in famTransmissionCols) {
      if(length(unique(vfData[,j]))==2) {
        tbl <- table(vfData[,i],vfData[,j])
        ft <- fisher.test(tbl)
        pValues <- append(pValues,ft$p.value)
        iVals <- append(iVals, i)
        jVals <- append(jVals, j)
        odds <- append(odds, ft$estimate)
        confIntLow <- append(confIntLow, ft$conf.int[1])
        confIntHigh <- append(confIntHigh, ft$conf.int[2])
      }
    }
  }
}

pValuesAdj <- p.adjust(pValues, method = 'BH')
names(pValuesAdj) <- pValues
pThreshold <- 0.05

data.frame(x=sort(pValues),
           y=sort(pValuesAdj)) %>%
  mutate(significance=if_else(x < pThreshold, "significant before correction", "not significant")) %>%
  mutate(significance=if_else(y < pThreshold, "significant", significance)) %>%
  ggplot(aes(x=x,y=y, color=significance)) +
  geom_point() +
  scale_color_manual(values=c("significant"="red","not significant"="black", "significant before correction"="ForestGreen")) +
  geom_hline(yintercept = pThreshold) + geom_vline(xintercept = pThreshold) +
  xlab("Significance ") + ylab("p-value")

data.frame(pval=1/pValuesAdj,
                 odds=odds,
                 cil=confIntLow,
                 cih=confIntHigh,
                 i=iVals,
                 j=jVals,
                 label=paste0(colnames(vfData)[iVals], " vs ",colnames(vfData)[jVals]),
                 stringsAsFactors = FALSE) %>%
        filter(pValuesAdj < 0.05) %>%
        mutate(maxcih=max(cih[is.finite(cih)])) %>%
        mutate(odds=if_else(is.finite(odds),odds,maxcih),
               cih=if_else(is.finite(cih),cih,maxcih)) %>%
        arrange(-pval, odds, -cih, cil) %>%
        arrange(pval, -odds, -cih, cil) %>%
        mutate(label=factor(label,levels=label)) %>%
        ggplot(aes(x=label,y=odds)) +
        geom_point(stat="identity") +
        geom_segment(aes(x=label,xend=label,y=cil,yend=cih)) +
        geom_hline(yintercept = 1, color="red") +
        ylab("Odds ratio (+ confidence interval)") + xlab("Test performed") +
        scale_y_log10() +
        coord_flip()
```

```{r fisher.tests.viralFamilies, echo=FALSE, eval=printTests, results='asis', fig.height=3}


for(i in famAttributeCols) {
  attribute <- colnames(vfData)[i]
  if(length(unique(vfData[,i]))==2) {
    cat("\n\n\\pagebreak\n")
    cat(paste0("\n\n# Comparison of attribute '", attribute, "' for viral families\n"))
    for(j in famTransmissionCols) {
      if(length(unique(vfData[,j]))==2) {
        transmission <- colnames(vfData)[j]
        cat(paste0("\n\n### Transmission mode '", transmission, "' vs attribute '", attribute, "'\n"))
        tbl <- table(vfData[,i],vfData[,j])
        df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
          mutate(Var1=paste0(attribute, " = ", Var1))
        colnames(df) <- c("",transmission,"other")
        print(kable(df))
        ft <- fisher.test(tbl)
        pValue <- ft$p.value
        pValueAdj <- pValuesAdj[as.character(pValue)]
        color <- if_else(pValueAdj < pThreshold,"red",if_else(pValue < pThreshold, "ForestGreen", "black"))
        cat(paste0("\\textcolor{",color,"}{p-value = ",round(pValue,digits = 5),"}\\newline"))
        cat(paste0("\\textcolor{",color,"}{adjusted p-value = ",round(pValueAdj,digits = 5),"}"))
      }
    }
  }
}

```

```{r figures.human.viralFamilies, echo=FALSE, eval=printTestFigures, results='asis', fig.height=3}

cat("\n\n\\pagebreak\n")
cat(paste0("\n\n# Analysis for viral families in humans\n"))


pValues <- NULL
odds <- NULL
confIntLow <- NULL
confIntHigh <- NULL
iVals <- NULL
jVals <- NULL

for(i in famAttributeColsH) {
  if(length(unique(vfDataH[,i]))==2) {
    for(j in famTransmissionColsH) {
      if(length(unique(vfDataH[,j]))==2) {
        tbl <- table(vfDataH[,i],vfDataH[,j])
        ft <- fisher.test(tbl)
        pValues <- append(pValues,ft$p.value)
        iVals <- append(iVals, i)
        jVals <- append(jVals, j)
        odds <- append(odds, ft$estimate)
        confIntLow <- append(confIntLow, ft$conf.int[1])
        confIntHigh <- append(confIntHigh, ft$conf.int[2])
      }
    }
  }
}

pValuesAdj <- p.adjust(pValues, method = 'BH')
names(pValuesAdj) <- pValues
pThreshold <- 0.05

data.frame(x=sort(pValues),
           y=sort(pValuesAdj)) %>%
  mutate(significance=if_else(x < pThreshold, "significant before correction", "not significant")) %>%
  mutate(significance=if_else(y < pThreshold, "significant", significance)) %>%
  ggplot(aes(x=x,y=y, color=significance)) +
  geom_point() +
  scale_color_manual(values=c("significant"="red","not significant"="black", "significant before correction"="ForestGreen")) +
  geom_hline(yintercept = pThreshold) + geom_vline(xintercept = pThreshold) +
  xlab("Significance ") + ylab("p-value")

data.frame(pval=1/pValuesAdj,
                 odds=odds,
                 cil=confIntLow,
                 cih=confIntHigh,
                 i=iVals,
                 j=jVals,
                 label=paste0(colnames(vfDataH)[iVals], " vs ",colnames(vfDataH)[jVals]),
                 stringsAsFactors = FALSE) %>%
        filter(pValuesAdj < 0.05) %>%
        mutate(maxcih=max(cih[is.finite(cih)])) %>%
        mutate(odds=if_else(is.finite(odds),odds,maxcih),
               cih=if_else(is.finite(cih),cih,maxcih)) %>%
        arrange(-pval, odds, -cih, cil) %>%
        arrange(pval, -odds, -cih, cil) %>%
        mutate(label=factor(label,levels=label)) %>%
        ggplot(aes(x=label,y=odds)) +
        geom_point(stat="identity") +
        geom_segment(aes(x=label,xend=label,y=cil,yend=cih)) +
        geom_hline(yintercept = 1, color="red") +
        ylab("Odds ratio (+ confidence interval)") + xlab("Test performed") +
        scale_y_log10() +
        coord_flip()

```

```{r fisher.tests.human.viralFamilies, echo=FALSE, eval=printTests, results='asis', fig.height=3}

for(i in famAttributeColsH) {
  attribute <- colnames(vfDataH)[i]
  if(length(unique(vfDataH[,i]))==2) {
    cat("\n\n\\pagebreak\n")
    cat(paste0("\n\n# Comparison of attribute '", attribute, "' for viral families in humans\n"))
    for(j in famTransmissionColsH) {
      if(length(unique(vfDataH[,j]))==2) {
        transmission <- colnames(vfDataH)[j]
        cat(paste0("\n\n### Transmission mode '", transmission, "' vs attribute '", attribute, "' in humans\n"))
        tbl <- table(vfDataH[,i],vfDataH[,j])
        df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
          mutate(Var1=paste0(attribute, " = ", Var1))
        colnames(df) <- c("",transmission,"other")
        print(kable(df))
        ft <- fisher.test(tbl)
        pValue <- ft$p.value
        pValueAdj <- pValuesAdj[as.character(pValue)]
        color <- if_else(pValueAdj < pThreshold,"red",if_else(pValue < pThreshold, "ForestGreen", "black"))
        cat(paste0("\\textcolor{",color,"}{p-value = ",round(pValue,digits = 5),"}\\newline"))
        cat(paste0("\\textcolor{",color,"}{adjusted p-value = ",round(pValueAdj,digits = 5),"}"))
      }
    }
  }
}


```

```{r figures.nonhuman.viralFamilies, echo=FALSE, eval=printTestFigures, results='asis', fig.height=3}

cat("\n\n\\pagebreak\n")
cat(paste0("\n\n# Analysis for viral families in non-human hosts\n"))


pValues <- NULL
odds <- NULL
confIntLow <- NULL
confIntHigh <- NULL
iVals <- NULL
jVals <- NULL

for(i in famAttributeCols) {
  if(length(unique(vfData[,i]))==2) {
    for(j in famTransmissionCols) {
      if(length(unique(vfData[,j]))==2) {
        tbl <- table(vfData[,i],vfData[,j])
        ft <- fisher.test(tbl)
        pValues <- append(pValues,ft$p.value)
        iVals <- append(iVals, i)
        jVals <- append(jVals, j)
        odds <- append(odds, ft$estimate)
        confIntLow <- append(confIntLow, ft$conf.int[1])
        confIntHigh <- append(confIntHigh, ft$conf.int[2])
      }
    }
  }
}

pValuesAdj <- p.adjust(pValues, method = 'BH')
names(pValuesAdj) <- pValues
pThreshold <- 0.05

data.frame(x=sort(pValues),
           y=sort(pValuesAdj)) %>%
  mutate(significance=if_else(x < pThreshold, "significant before correction", "not significant")) %>%
  mutate(significance=if_else(y < pThreshold, "significant", significance)) %>%
  ggplot(aes(x=x,y=y, color=significance)) +
  geom_point() +
  scale_color_manual(values=c("significant"="red","not significant"="black", "significant before correction"="ForestGreen")) +
  geom_hline(yintercept = pThreshold) + geom_vline(xintercept = pThreshold) +
  xlab("Significance ") + ylab("p-value")

data.frame(pval=1/pValuesAdj,
                 odds=odds,
                 cil=confIntLow,
                 cih=confIntHigh,
                 i=iVals,
                 j=jVals,
                 label=paste0(colnames(vfData)[iVals], " vs ",colnames(vfData)[jVals]),
                 stringsAsFactors = FALSE) %>%
        filter(pValuesAdj < 0.05) %>%
        mutate(maxcih=max(cih[is.finite(cih)])) %>%
        mutate(odds=if_else(is.finite(odds),odds,maxcih),
               cih=if_else(is.finite(cih),cih,maxcih)) %>%
        arrange(-pval, odds, -cih, cil) %>%
        arrange(pval, -odds, -cih, cil) %>%
        mutate(label=factor(label,levels=label)) %>%
        ggplot(aes(x=label,y=odds)) +
        geom_point(stat="identity") +
        geom_segment(aes(x=label,xend=label,y=cil,yend=cih)) +
        geom_hline(yintercept = 1, color="red") +
        ylab("Odds ratio (+ confidence interval)") + xlab("Test performed") +
        scale_y_log10() +
        coord_flip()

```

```{r fisher.tests.nonhuman.viralFamilies, echo=FALSE, eval=printTests, results='asis', fig.height=3}

for(i in famAttributeCols) {
  attribute <- colnames(vfData)[i]
  if(length(unique(vfData[,i]))==2) {
    cat("\n\n\\pagebreak\n")
    cat(paste0("\n\n# Comparison of attribute '", attribute, "' for viral families in non-human hosts\n"))
    for(j in famTransmissionCols) {
      if(length(unique(vfData[,j]))==2) {
        transmission <- colnames(vfData)[j]
        cat(paste0("\n\n### Transmission mode '", transmission, "' vs attribute '", attribute, "' in non-human hosts\n"))
        tbl <- table(vfData[,i],vfData[,j])
        df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
          mutate(Var1=paste0(attribute, " = ", Var1))
        colnames(df) <- c("",transmission,"other")
        print(kable(df))
        ft <- fisher.test(tbl)
        pValue <- ft$p.value
        pValueAdj <- pValuesAdj[as.character(pValue)]
        color <- if_else(pValueAdj < pThreshold,"red",if_else(pValue < pThreshold, "ForestGreen", "black"))
        cat(paste0("\\textcolor{",color,"}{p-value = ",round(pValue,digits = 5),"}\\newline"))
        cat(paste0("\\textcolor{",color,"}{adjusted p-value = ",round(pValueAdj,digits = 5),"}"))
      }
    }
  }
}
```

```{r viralFamiliesByHost, echo=FALSE, eval=printByHost, results='asis', fig.height=3}

for(host in unique(vfData$Common.Name)) {
  cat("\n\n\\pagebreak\n")
  cat(paste0("\n\n# Analysis for host '", host, "' for viral families\n"))
  
  hostvfData <- data %>%
  group_by(Host.species, Common.Name, Virus.family, NucAcid) %>%
  summarize_if(is.logical, function(X) { if(sum(X) > 0) { return(TRUE) } else { return(FALSE)}}) %>%
  ungroup() %>%
  as.data.frame() %>%
  filter(Common.Name==host)
  
  pValues <- NULL
  odds <- NULL
  confIntLow <- NULL
  confIntHigh <- NULL
  iVals <- NULL
  jVals <- NULL
  
  for(i in famAttributeCols) {
    if(length(unique(hostvfData[,i]))==2) {
      for(j in famTransmissionCols) {
        if(length(unique(hostvfData[,j]))==2) {
          tbl <- table(hostvfData[,i],hostvfData[,j])
          ft <- fisher.test(tbl)
          pValues <- append(pValues,ft$p.value)
          iVals <- append(iVals, i)
          jVals <- append(jVals, j)
          odds <- append(odds, ft$estimate)
          confIntLow <- append(confIntLow, ft$conf.int[1])
          confIntHigh <- append(confIntHigh, ft$conf.int[2])
        }
      }
    }
  }
  
  pValuesAdj <- p.adjust(pValues, method = 'BH')
  names(pValuesAdj) <- pValues
  pThreshold <- 0.05
  
  if(printTestFigures) {
    print(data.frame(x=sort(pValues),
               y=sort(pValuesAdj)) %>%
      mutate(significance=if_else(x < pThreshold, "significant before correction", "not significant")) %>%
      mutate(significance=if_else(y < pThreshold, "significant", significance)) %>%
      ggplot(aes(x=x,y=y, color=significance)) +
      geom_point() +
      scale_color_manual(values=c("significant"="red","not significant"="black", "significant before correction"="ForestGreen")) +
      geom_hline(yintercept = pThreshold) + geom_vline(xintercept = pThreshold) +
      xlab("Significance ") + ylab("p-value"))
  
    print(data.frame(pval=1/pValuesAdj,
                   odds=odds,
                   cil=confIntLow,
                   cih=confIntHigh,
                   i=iVals,
                   j=jVals,
                   label=paste0(colnames(hostvfData)[iVals], " vs ",colnames(hostvfData)[jVals]),
                   stringsAsFactors = FALSE) %>%
          filter(pValuesAdj < 0.05) %>%
          mutate(maxcih=max(cih[is.finite(cih)])) %>%
          mutate(odds=if_else(is.finite(odds),odds,maxcih),
                 cih=if_else(is.finite(cih),cih,maxcih)) %>%
          arrange(-pval, odds, -cih, cil) %>%
          arrange(pval, -odds, -cih, cil) %>%
          mutate(label=factor(label,levels=label)) %>%
          ggplot(aes(x=label,y=odds)) +
          geom_point(stat="identity") +
          geom_segment(aes(x=label,xend=label,y=cil,yend=cih)) +
          geom_hline(yintercept = 1, color="red") +
          ylab("Odds ratio (+ confidence interval)") + xlab("Test performed") +
          scale_y_log10() +
          coord_flip())
  }
  
  if(printTests) {
    for(i in famAttributeCols) {
      attribute <- colnames(hostvfData)[i]
      cat("\n\n\\pagebreak\n")
      if(length(unique(hostvfData[,i]))==1) {
        next
      }
      cat(paste0("\n\n# Comparison of attribute '", attribute, "' for host ", host, " for viral families"))
      for(j in famTransmissionCols) {
        if(length(unique(hostvfData[,j]))==1) {
          next
        }
        transmission <- colnames(hostvfData)[j]
        cat(paste0("\n\n### Transmission mode '", transmission, "' vs attribute '", attribute, "' in ", host, "\n"))
        tbl <- table(hostvfData[,i],hostvfData[,j])
        df <- as.data.frame(tbl) %>% spread(Var2, Freq) %>% select(1,3,2) %>%
          mutate(Var1=paste0(attribute, " = ", Var1))
        colnames(df) <- c("",transmission,"other")
        print(kable(df))
        ft <- fisher.test(tbl)
        pValue <- ft$p.value
        pValueAdj <- pValuesAdj[as.character(pValue)]
        color <- if_else(pValueAdj < pThreshold,"red",if_else(pValue < pThreshold, "ForestGreen", "black"))
        cat(paste0("\\textcolor{",color,"}{p-value = ",round(pValue,digits = 5),"}\\newline"))
        cat(paste0("\\textcolor{",color,"}{adjusted p-value = ",round(pValueAdj,digits = 5),"}"))
      }
    }
  }
}

```


```{r heatmap, echo=FALSE, eval=printHeatmap, results='asis', fig.height=9}

VariableLevels <- c(attributeVars, transmissionVars, "Baltimore.classification")
VariableCols <- which(colnames(data) %in% VariableLevels)

plotData <- data %>% select(Virus.name, Virus.family, Virus.genus, VariableCols) %>%
  gather("Variable","Value", VariableLevels) %>%
  mutate(subject=ifelse(Variable=="Lipid.Envelope","Lipid",
                        ifelse(Variable %in% c("Spherical","Bullet.form","Filamentous", "Pleomorphic"),
                               "Capsid Structure",
                               ifelse(Variable %in% c("Baltimore.classification","Category"),"other",
                                      ifelse(Variable=="NucAcid","NucAcid","Transmission mode")))),
         color=ifelse(subject=="Lipid" & Value==TRUE,"red",
                      ifelse(subject=="Capsid Structure" & Value==TRUE, "blue",
                             ifelse(subject=="Transmission mode" & Value==TRUE,"green",
                                    ifelse(Value=="DNA","cyan",
                                           ifelse(Value=="RNA","yellow",
                                                  ifelse(subject=="other","white","light grey")))))),
         labelText=ifelse(subject %in% c("other","NucAcid"),Value,"")) %>%
  mutate(species=factor(Virus.name, levels = sort(unique(data$Virus.name), decreasing = TRUE))) %>%
  mutate(subject=factor(subject, levels = c("other","Capsid Structure","NucAcid","Lipid","Transmission mode"))) %>%
  mutate(Variable=factor(Variable, levels = VariableLevels))

plotFamilyOrder <- plotData %>% filter(Variable=="Baltimore.classification") %>% arrange(Value, Virus.family) %>% pull(Virus.family) %>% unique()

plotData %>%
  mutate(Virus.family=factor(Virus.family, levels=plotFamilyOrder)) %>%
  ggplot(aes(x=Variable,y=species, fill=color)) +
  geom_tile(color="black") +
  geom_text(aes(label = labelText), size = 1.5) +
  scale_fill_manual(values = c("red"="red", "blue"="blue","green"="green","light grey"="light grey","white"="white","cyan"="cyan","yellow"="yellow")) +
  xlab("") + ylab("") +
  facet_grid(Virus.family ~ subject, scales = "free", space = "free", switch="y") +
  theme(strip.text.y = element_text(angle=180, size=6, margin = margin(0,0,0,0, "cm")),
        strip.placement = "outside",
        axis.text.y = element_text(size=6),
        strip.text.x = element_blank(),
        axis.text.x = element_text(size=6, angle=45, vjust = 1, hjust = 1),
        panel.spacing.y = unit(0.025, "lines"),
        panel.spacing.x = unit(0.025, "lines"),
        panel.background = element_rect(color = "black", fill = "black", size = 10),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

page1FamilyList <- plotFamilyOrder[1:ceiling(length(plotFamilyOrder)/2)]

plotData %>% filter(Virus.family %in% page1FamilyList) %>%
  mutate(Virus.family=factor(Virus.family, levels=plotFamilyOrder)) %>%
  ggplot(aes(x=Variable,y=species, fill=color)) +
  geom_tile(color="black") +
  geom_text(aes(label = labelText), size = 1.5) +
  scale_fill_manual(values = c("red"="red", "blue"="blue","green"="green","light grey"="light grey","white"="white","cyan"="cyan","yellow"="yellow")) +
  xlab("") + ylab("") +
  facet_grid(Virus.family ~ subject, scales = "free", space = "free", switch="y") +
  theme(strip.text.y = element_text(angle=180, size=6, margin = margin(0,0,0,0, "cm")),
        strip.placement = "outside",
        axis.text.y = element_text(size=6),
        strip.text.x = element_blank(),
        axis.text.x = element_text(size=6, angle=45, vjust = 1, hjust = 1),
        panel.spacing.y = unit(0.025, "lines"),
        panel.spacing.x = unit(0.025, "lines"),
        panel.background = element_rect(color = "black", fill = "black", size = 10),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

plotData %>% filter(Virus.family %!in% page1FamilyList) %>%
  mutate(Virus.family=factor(Virus.family, levels=plotFamilyOrder)) %>%
  ggplot(aes(x=Variable,y=species, fill=color)) +
  geom_tile(color="black") +
  geom_text(aes(label = labelText), size = 1.5) +
  scale_fill_manual(values = c("red"="red", "blue"="blue","green"="green","light grey"="light grey","white"="white","cyan"="cyan","yellow"="yellow")) +
  xlab("") + ylab("") +
  facet_grid(Virus.family ~ subject, scales = "free", space = "free", switch="y") +
  theme(strip.text.y = element_text(angle=180, size=6, margin = margin(0,0,0,0, "cm")),
        strip.placement = "outside",
        axis.text.y = element_text(size=6),
        strip.text.x = element_blank(),
        axis.text.x = element_text(size=6, angle=45, vjust = 1, hjust = 1),
        panel.spacing.y = unit(0.025, "lines"),
        panel.spacing.x = unit(0.025, "lines"),
        panel.background = element_rect(color = "black", fill = "black", size = 10),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())


```

```{r humanHeatmap, echo=FALSE, eval=printHumanHeatmap, results='asis', fig.height=9}

VariableLevels <- c(attributeVars, transmissionVars, "Baltimore.classification")
VariableCols <- which(colnames(data) %in% VariableLevels)

humanPlotData <- data %>%
  filter(Common.Name=="Human") %>%
  select(Virus.name, Virus.family, Virus.genus, VariableCols) %>%
  gather("Variable","Value", VariableLevels) %>%
  mutate(subject=ifelse(Variable=="Lipid.Envelope","Lipid",
                        ifelse(Variable %in% c("Spherical","Bullet.form","Filamentous", "Pleomorphic"),
                               "Capsid Structure",
                               ifelse(Variable %in% c("Baltimore.classification","Category"),"other",
                                      ifelse(Variable=="NucAcid","NucAcid","Transmission mode")))),
         color=ifelse(subject=="Lipid" & Value==TRUE,"red",
                      ifelse(subject=="Capsid Structure" & Value==TRUE, "blue",
                             ifelse(subject=="Transmission mode" & Value==TRUE,"green",
                                    ifelse(Value=="DNA","cyan",
                                           ifelse(Value=="RNA","yellow",
                                                  ifelse(subject=="other","white","black")))))),
         labelText=ifelse(subject %in% c("other","NucAcid"),Value,"")) %>%
  mutate(species=factor(Virus.name, levels = sort(unique(data$Virus.name), decreasing = TRUE))) %>%
  mutate(subject=factor(subject, levels = c("other","Capsid Structure","NucAcid","Lipid","Transmission mode"))) %>%
  mutate(Variable=factor(Variable, levels = VariableLevels))

humanPlotFamilyOrder <- humanPlotData %>% filter(Variable=="Baltimore.classification") %>% arrange(Value, Virus.family) %>% pull(Virus.family) %>% unique()

humanPlotData %>%
  mutate(Virus.family=factor(Virus.family, levels=humanPlotFamilyOrder)) %>%
  ggplot(aes(x=Variable,y=species, fill=color)) +
  geom_tile(aes(width=1),color="black") +
  geom_text(aes(label = labelText), size = 1.5) +
  scale_fill_manual(values = c("red"="red", "blue"="blue","green"="green","light grey"="light grey","white"="white","cyan"="cyan","yellow"="yellow", "black"="black")) +
  xlab("") + ylab("") +
  facet_grid(Virus.family ~ subject, scales = "free", space = "free", switch="y") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(size=6, angle=45, vjust = 1, hjust = 1),
        strip.text.y = element_text(angle=180, size=5, margin = margin(0,0,0,0, "cm")),
        strip.placement = "outside",
        strip.text.x = element_blank(),
        panel.spacing.y = unit(0.15, "lines"),
        panel.spacing.x = unit(-0.3, "lines"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none")

humanPlotData %>%
  mutate(Virus.family=factor(Virus.family, levels=humanPlotFamilyOrder)) %>%
  ggplot(aes(x=Variable,y=species, fill=color)) +
  geom_tile(aes(width=1),color="black") +
  geom_text(aes(label = labelText), size = 1.5) +
  scale_fill_manual(values = c("red"="red", "blue"="blue","green"="green","light grey"="light grey","white"="white","cyan"="cyan","yellow"="yellow", "black"="black")) +
  xlab("") + ylab("") +
  facet_grid(Virus.family ~ subject, scales = "free", space = "free", switch="y") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(size=6, angle=45, vjust = 1, hjust = 1),
        strip.text.y = element_text(angle=180, size=4, margin = margin(0,0,0,0, "cm")),
        strip.placement = "outside",
        strip.text.x = element_blank(),
        strip.background=element_rect(color="black",size=0.1),
        panel.spacing.y = unit(-0.1, "lines"),
        panel.spacing.x = unit(-0.3, "lines"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none")


```
